A forest is more than its trees: haplotypes and ancestral recombination graphs

### 2025-05-25 paper/review-responses.tex

\begin{point}{}
 I really enjoyed the title of the paper!
\end{point}

\reply{
Thank you!
}

\begin{point}{\revref}
 I am a bit confused by the definition of weighted Robinson-Foulds distance at \revref.% on the bottom of page 6. 
I believe that this is more like a Robinson-Foulds \emph{similarity}, as I believe that as defined it should be maximized when the two ARGs are the same, and smaller for more dissimilar ARGs.
\end{point}

\reply{
    Oh dear, that was an embarrassing error
    (that happily doesn't affect anything else in the paper).
    This has been corrected.
}

@@

\begin{point}{}
 Is the dissimilarity measure defined in equation \eqref{eqn:wrmse} used anywhere?
\end{point}

\reply{No, the weighted mean square error defined in \eqref{eqn:wrmse} is not shown for our later experiments shown in \figref{fig:dissimilarity}, however it is always computed when using our implementation is \texttt{tscompare};
we now explicitly say this \revref.
}

@@

\begin{point}{}
 What is ``dissimilarity'' as mentioned in the caption of \figref{fig:conceptual_discrepancy}?
\end{point}

\reply{``dissimilarity'' was our original language to describe the non-matched spans, and we missed this use of the old name. We have now changed the caption in \figref{fig:conceptual_discrepancy} to reflect that change.
}

@@

\begin{point}{\revref}
 I found the discussion around the metric version to be hard to follow and worth expanding.
For example, I believe that the edges of the ARG are what's getting embedded into
$[0, L) \times N \times N$, and the symmetric differences are taken in this space in the obvious way (i.e., without this mapping its unclear what the relative difference of two ARGs is).
There's also a mention of ``dissimilarity'' in this paragraph. I assume that that's the ``symmetrized version of non-matched span'' that was just defined, but it would be good to make that clear. 
\end{point}

\reply{
    Thanks again to the reviewer for carefully thinking about what we wrote,
    which on further examination has turned out to be wrong in several ways (apologies!).
    Please see the updated version that we believe to be correct
    and thus hopefully less confusing. \revref
}

@@

\begin{point}{}
 The performance on inferred ARGs makes clear that this is a difficult problem, and that this haplotype view might be useful conceptually and for methods development, but that as of now we should probably not take these ancestral haplotypes at face value in inferred ARGs. It would be helpful if the authors could provide some intuition for what's going wrong in this case. Presumably the topologies are not inferred well, and many nodes can be extended, which often results in some true and false positives? Another use case I could see would be using this approach to simplify inferred ARGs to make various downstream computations faster (a la \figref{fig:speed_and_edges}). As such, I would be curious to see how the extension algorithm affects the number of edges in inferred ARGs.
\end{point}

\reply{
    We agree on all counts -- in particular,
    it would be helpful to dig into why haplotypes are being inferred wrongly;
    however, this more detailed investigation is beyond the scope of this paper
    (since it's more an investigation into the inference methods).
    More generally: if an inferred ARG is incorrect for certain nodes, our method of extend haplotypes exacerbates this issue as extending an incorrectly inferred node extends this incorrect node span even further; we've added a note to this effect \revref.
    \comment{TODO: add number of edges?}
    Based on our tests in \figref{fig:speed_and_edges} and \tabref{tab:edge-counts}, these results would additionally hold for the inferred ARGs.
    For this reason, we didn't include additional figures for edge counts and computation time on inferred ARGs.
}

@@

\begin{point}{}
 This is obviously a matter of taste, but I would have found the paper a bit easier to follow if some of the points discussed in the discussion were brought up earlier in the paper. For example, the ``ignorance and omission in the ARG'' section \llname{ignorance}
answered questions I had when reading the bits around \figref{fig:conceptual} -- upon initially reading the bit around \figref{fig:conceptual}
I was confused by the interpretation that the lack of a node implies that that individual is not represented in other parts of the ARG (as any point along any branch in an ARG represents \emph{some} individual ancestral to sample), and as a result I found the presented interpretation somewhat unnatural.
This was then eventually clarified (very eloquently) in the discussion.
Similarly, I would have appreciated the connection to parsimony earlier.
It was not completely obvious to me that recovering the unary nodes from the succinct tree sequence might not be possible in all cases, and as such that the algorithm is an attempt at inference, and furthermore, that it's optimizing some reasonable objective. Again, this point was addressed very well, but I had to wait until the discussion.
\end{point}

\reply{
    This is a good suggestion;
    rather than break up the flow (for less sophisticated readers)
    we've inserted a pointer to these parts of the discussion, at \revref.
}

@@

\begin{point}{\revref} % p.8 l.25
 ``we simulate ARGs containing full haplotypes using msprime'': since msprime can simulate ARGs under various models, please specify here which model is used (SMC, SMC'... ?).
\end{point}

\reply{
Good suggestion; we've done this (we used ``hudson''). \revref
}

\begin{point}{}
In general, it comes natural to wonder whether the algorithms can be corrected to avoid incorrectly added spans
(e.g. \revref). % p.10 l. 5).
It would be worth to mention whether an attempt of correction has been made and to explain the difficulties/limitations involved.
\end{point}

\reply{
    Good suggestion; we've added a note about this. \llname{bad_spans}
}

@@

\reply{Now updated. \revref
}

%%%%%%%%%%%%%%
\reviewersection{3}

\begin{quote}
The fundamental idea of this paper is important and the worked-out
algorithm appears quite valuable. I want to see it published. However
the current version is so difficult to read that it is likely to
have a very limited audience, and appears to have some errors.
\end{quote}

We're glad you think it's important, and apologies for the errors.

@@

\begin{quote}
I do not feel I have understood the entire paper despite several
hours' work. Some of my comments may be misunderstandings, but if
so, the text I misunderstood could probably use clarification.
\end{quote}

As you'll see, we've tried to improve clarity throughout;
hopefully it is more digestable now.

@@

\begin{point}{} % (1)
The description of \algorithmref{alg:extend} and \algorithmref{alg:mergeable} is confusing.
Naming them this way implies that they are separate algorithms but
in fact ``Algorithm 2'' is the definition of a function needed by
Algorithm 1.
\end{point}

\reply{
    We think it's reasonably standard for one algorithm to make use of another
    (for instance, many algorithms apply a \texttt{sort} at some point).
    However, it's a good point that referring abstractly to ``\algorithmref{alg:extend}'' and 
    ``\algorithmref{alg:mergeable}'' is a little abstract and hard to remember.
    So, we've gone back to places where the text just said ``\algorithmref{alg:extend}'',
    and where we thought helpful, changed it to something like
    ``extending haplotypes with \algorithmref{alg:extend}''.
    For more background:
    originally we had both \algorithmref{alg:extend} and \algorithmref{alg:mergeable} within 
	one algorithm environment.
    However, the \texttt{Mergeable} is a conceptually separate operation
    that by separating it out makes understanding \texttt{ExtendHaplotypes} easier.
	\algorithmref{alg:extend} is the first algorithm as it is the main construction and the most important of the two.
}

\begin{point}{} %(2)
\algorithmref{alg:mergeable} is mentioned in an explanation a full page before it
is introduced.
\end{point}

\reply{
    This is a consequence of the typesetting: \algorithmref{alg:mergeable} 
    is in a floating environment, and
    the earliest the environment will appear is the next page following the previous figure.
    Rest assured that Genetics will typeset this appropriately.
}

\begin{point}{} %(3)
\algorithmref{alg:extend} seems to be named ExtendHaplotypes and \algorithmref{alg:mergeable} seems
to be named Mergeable: having two names for each one is not helpful.
\end{point}

\reply{
    This point overlaps with point 3.1 -- see our response above for how we've hopefully
    improved the situation.
    We like to think of ExtendHaplotypes and Mergable as the names of the functions contained in \algorithmref{alg:extend} and \algorithmref{alg:mergeable}, respectively.
}

\begin{point}{} %(4)
``Mergeable'' is an unfortunate name for a function which does not
return a yes or no answer but instead returns the number of
new edges to be added. Also the text needs a comment about why
infinity is sometimes returned.
\end{point}

\reply{
    A reasonable point; we've changed the name to \texttt{MergeNum}.
	For clarity, we have also included a note about returning infinity in the description of \algorithmref{alg:mergeable}.
}

\begin{point}{} %(5)
In \algorithmref{alg:extend} the variable $m$ means the number of new edges to be
added. In \algorithmref{alg:mergeable} that quantity is called $m_e$ and $m$ means the
number of extended haplotypes. This is confusing. Please use the
same name for a given concept in all pseudocode.
\end{point}

\reply{Good point, this change has been made.
}

\begin{point}{} %(6)
Comments in the pseudocode would make it easier to follow, as could
use of more meaningful variable names (I particularly dislike $M$, $m$, $M'$
in the same algorithm).
\end{point}

\reply{Good point. We have included some comments for clarity.
However, we're going to respectfully differ on stylistic choice of concise variable names.
}

@@

\begin{point}{}
Capital $I$ is an unfortunate subscript as it is almost indistinguishable
from digit 1.
\end{point}

\reply{While we agree in general,
within the context of our algorithm we are interested paths of inheritance 
	between the edges removed from a local tree (the out-forest) and the edges just added into the next local tree (the in-forest). 
	Because of our language use of out and in, we feel like our choice of $O$ and $I$ is reasonable.
}

@@

\begin{quote} % (2)
\textbf{Issues in algorithms.}
\end{quote}

\begin{point}{}
\algorithmref{alg:extend} appears to have an error leading to an infinite
loop. Variable $M$ is initially set to 0. It is compared to variable $m$,
which is a number of new edges to be added and must be positive or zero.
Thus, $m$ will never be lower than $M$, so the action taken is to set $M'$ to
the minimum of $m$ and $M$, which will always be zero. $M$ is then set to
$M'$, that is, to zero. Since this is inside a while loop with terminating
condition $M$ == infinity, it does not terminate. (I coded it to check.)

The logic problem is probably around line 10. We have just tested
whether $m < M$ and this was false, so $m \ge M$. It is therefore
strange to take the minimum of $m$ and $M,$ as this will always be
$M$. However I can't work out the correct code.
\end{point}

\reply{
Good catch -- and, many apologies --
that is indeed a bug; the correction changes line 7 to
``if $m \le M$'' and line 10 to ``Set $M' = \min(m_e, M')$.''.
(To be clear, the bug was in the paper, not the implemented code;
if the reviewer is interested they can see our implementation
in the tskit library at
\href{https://github.com/tskit-dev/tskit/blob/c6b06b0c4f90b5098aedddca7dd0e3fbbad964d0/c/tskit/trees.c#L9078}{c/tskit/trees.c:9078} in the current release.)
}

@@

\begin{point}{}
The text description of \algorithmref{alg:extend} (ExtendHaplotypes) says that
paths with the fewest new edges are added first. The pseudocode
given does not appear to do that. It appears (bearing in mind that
it has at least one bug, so I can't be positive) to be willing to
execute a Merge before examining all children, so I think it will
Merge the first child with a better score than previous children,
regardless of the fact that an even better child may remain to be
found.
\end{point}

\reply{
We apologize again for the bug in the previous point that caused this misunderstanding.
Fixed, it should perform as advertised
(and, we've added a bit clarifying why it works
to the description of \algorithmref{alg:extend}).
}

\begin{point}{}
After considerable work I have been unable to code \algorithmref{alg:mergeable}.
When is a node NULL? Does this mean the node doesn't exist (for
example, the parent of the root)?
\end{point}

\reply{
    A node is NULL when it has no parent; this is now explained in the description
    of the algorithm.
    Apologies, this is part of the conventions in tskit.
}

@@

\begin{point}{} % (3)
I recommend that the example partial tree in \figref{fig:conceptual} be
extended to a full tree (it only needs one more node) and used in
the practical examples in the text.
\end{point}

\reply{
Good idea,
we now reference the figure in the text,
at \revref{} and \llname{ref_conceptual2}.
On further consideration we think that adding the rest of the tree
here would make the figure too confusing
(not so much for the trees in A\&C, but more so in the haplotype diagrams B\&D).
}

@@

\begin{point}{} % (4) 
The concrete example under ``An algorithm to extend haplotypes'' is
difficult to visualize. Can it be related to \figref{fig:conceptual}? When I try to
do this, though, I have difficulty because in \figref{fig:conceptual} the right-hand
tree does not \emph{have} $v_1 \ldots v_n$, it jumps straight from $p$ to $c.$ Is
this example not mergeable, or is the description of mergeablility
missing any indication that some of the nodes it relies on may not
exist?
\end{point}

\reply{Good point. We have connected our concrete construction in that section
	 to reference \figref{fig:conceptual} to improve clarity. \revref
}

\begin{point}{} % (5) 
The direction in which time is measured on the tree needs to be stated
as unfortunately the literature has both conventions. Also, the text
implies at several points that this can be used for trees without
time information, but it cannot as times are used in the algorithm.
\end{point}

\reply{We state how time is measured here \revref;
we've added a reminder at \llname{time_reminder}.
We do use time in our algorithm, but mainly as a way to break ties, as is the case with \eqnref{eqn:matched-spans}. This tie break will default to the first node if the ARG does not contain time information.
We've also added a note about undated ARGs \llname{undated}.
}

@@

\begin{point}{}
Following on from this, in \algorithmref{alg:extend}, why can no node in $u$ be at the
same time as a node in $v?$ At first I thought we were assuming that no
two nodes, other than samples, are ever at the same time (a fairly
common assumption) but then we would not need this rule. Is this just
a convenience to avoid putting two nodes on the same branch at the same
timepoint? What happens if this algorithm is attempted with inferred
ARGs that discretize time, as some inference algorithms do, so that ties
become common?
\end{point}

\reply{
    This is exactly right:
    we don't assume that node times are unique,
    as it's common (e.g., from discrete-time simulators)
    for this not to be true;
    so the check is to avoid producing an illegal situation
    (parent and child at the same time).
    We now explain this at \llname{undated}.
}

\begin{point}{\llname{truth}}
``In what follows we think of $T_2$ as the true ARG and $T_1$ as an
inferred ARG.'' This is not strongly enough stated. The measure is
asymmetrical with regards to $T_1$ and $T_2$, and will give wrong answers
if they are reversed. It is also not suitable for comparing two
ARGs where neither is the true ARG.
\end{point}

\reply{
    We've added extra emphasis to the point that the quantities are not symmetric
    \llname{asymmetric};
    furthermore the additional discussion in the ``Metrics on ARGs'' section
    should help reinforce this point.
    As for reversing $\T_1$ and $\T_2$,
	we think that while $\match(\T_2,\T_1)$ might not be the most useful quantity,
    it can still give some valuable information between inferred $\T_1$ and true $\T_2$.
    We discuss this when comparing inverse matched spans between $(\T_1,\T_2)$ and matched spans between $(\T_2, \T_1)$ \revref.
}

@@

\begin{point}{}
In general the section on comparison of true and inferred ARGs is
difficult to follow. \figref{fig:conceptual_discrepancy} helps but appears too late in the
paper.
\end{point}

\reply{
    We have expanded our example to hopefully give more clarity (see additional text
    in caption).
    The first reference to \figref{fig:conceptual_discrepancy} occurs at the end
    of the second paragraph in the section on comparison;
    it would be hard to include the example earlier
    (and, we're not sure how helpful it will be until
    $\arf$ and $\tpr$ are defined).
}

@@

\begin{point}{} % (7) 
In \figref{fig:conceptual_discrepancy} it is unfortunate to use to use an example where
the forward and inverse matches are the same. I would also appreciate
a brief explanation of why it's 46 and 47 in the denominator--I figured
it out but it's less than obvious. Why do sample nodes count, as they
cannot vary among trees? Doesn't this just add bookkeeping effort?
It's otherwise a very helpful figure though.
\end{point}

\reply{
    We sympathize with the reviewer that working through even a simple example like this
    is tricky and confusing (but, we also find it very helpful).
    Good point about the same forward and inverse matches.
    Note that we did construct this example so that matching for node 4 requires a max to be taken over shared spans of its two possible matches 4 and 9.
To show some variety, we have added computation of $\match(\T_2,\T_1)$ to show some additional variety and hopefully give more clarity to the methods themselves (see caption).
The samples are included because in practice they do \emph{not} necessarily always match:
consider a case where some of the samples are ancestors of others
(ancient samples; trios; etcetera).
}

@@

\begin{point}{} % (8) 
I cannot clearly distinguish the three purplish colors in \figref{fig:speed_and_edges}.
\end{point}

\reply{Thank you for this feedback. We have changed \figref{fig:speed_and_edges} and the subsequent supplementary figures (\figref{fig:speed_and_edges_constant}, \figref{fig:speed_and_edges_other}, \figref{fig:speed_and_edges_other_constant}) to have more color variety.
}

\begin{point}{} % (9) 
The interpretation of \figref{fig:node-spans} is completely unclear to me. I
think it needs more explanatory text. I have no idea what point is being
made by these diagrams.
\end{point}

\reply{Thanks for the feedback. We have added substantial additional explanation at \revref{} and the caption of \figref{fig:node-spans}.
}

### 2025-05-25 paper/writeup.tex
pass through reviews

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}}
\newcommand{\ind}{\mathbf{1}}
\newcommand{\tn}{\textnormal}
\newcommand{\ov}{\overline}
\newcommand{\tskit}{\texttt{tskit}}
\newcommand{\tscompare}{\texttt{tscompare}}
\newcommand{\tsinfer}{\texttt{tsinfer}}
\newcommand{\tsdate}{\texttt{tsdate}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\argmax}{\operatorname{argmax}}
\newcommand{\dis}{\operatorname{dis}}
\newcommand{\match}{\underset{\longrightarrow}{\operatorname{match}}}
\newcommand{\invmatch}{\underset{\longleftarrow}{\operatorname{match}}}
\newcommand{\nmatch}{\underset{\centernot\longrightarrow}{\operatorname{match}}}
\newcommand{\comment}[1]{{\color{violet} \it #1}}
\newcommand{\pluseq}{\mathrel{{+}{=}}}
\newcommand{\arf}\operatorname{ARF}
\newcommand{\tpr}\operatorname{TPR}

@@

So, given the ARG shown in \figref[A\&B]{fig:conceptual},
it should be possible to extend the ancestral haplotype represented by node 3
to obtain the ARG shown in \figref[C\&D]{fig:conceptual},
thus adding additional information to the ARG.
This might be surprising,
as intuition from phylogenetics suggests we can only infer
information about the branching points in the tree, not intermediate (unary) nodes.
The goal of this paper is to answer:
How can we do this, and how accurate is the resulting inference?

@@

We work with the \emph{succinct tree sequence} representation of ARGs (henceforth, ``tree sequence''),
to take advantage of the % extensive, stable, and well-documented set of open-source 
tools available in \tskit{} \citep{tskit}, 
and our terminology and notation follows \citet{ralph2020efficiently}.
For our purposes here,
a tree sequence $\T = (N, E)$ contains a set of \emph{nodes} $N$ 
which represent ancestral segments of genome,
and \emph{edges} $E$ which represent relationships between nodes over different regions of the genome.
Each node $n \in N$ has a \emph{time} $t_n$,
which is the amount of time in the past that the individual who carried that segment of genome lived. \revpoint{3}{13}
Some nodes are \emph{samples}, meaning that they represent genome sequences available as data.
% Typically, samples are exclusively leaves of $\T$, however it is not the case that all samples are leaves.
Each edge $e \in E$ describes inheritance between ancestors and their descendants within the population.
The ancestor and descendant are represented by the parent node $p_e$ and child node $c_e$ respectively \revpoint{2}{1}
over a segment of genome $[\ell_e, r_e)$.
% (The tree sequence data structure typically includes substantially more information, including genotypes, that we will not use in this paper.)
Suppose that the unique elements of the set of left and right edge endpoints
are $0 = a_0 < a_1 < \cdots < a_{n} = L$, where $L$ is the length of the genome.
Using this information, one can construct the sequence of 
trees $\left(T_1,...,T_{n}\right)$ that describe how the nodes are related to each other along the genome:
each $T_k$ is a tree whose nodes are in $N$
and that describes relationships on the half-open interval $[a_{k-1}, a_k)$.
Nodes represent (portions of) ancestral haplotypes, and so
we will use the terms interchangeably.
Not all nodes appear in each tree,
and we say $n \in T_k$ for a node $n$ if the tree $T_k$ describes at least one parent-child relationship
for node $n$.

@@

% JK note: earlier version said "relatively" - relative to what?
An efficient algorithm to do this is described in \algorithmref{alg:extend}.
% and implemented in \tskit{} as \texttt{extend\_haplotypes}.
The algorithm considers each tree transition from $T_k$ to $T_{k+1}$ in turn, updating its internal state
(which includes possibly modifying $T_k$) as it goes.
Suppose we are at the transition from tree $T_k$ to tree $T_{k+1}$,
which is done by first removing a set of edges $O$
and then adding another set of edges $I$.
$O$ defines a sub-forest $F_O$ of $T_k$,
and $I$ defines a sub-forest $F_I$ of $T_{k+1}$.
In \figref{fig:conceptual} (in the portion of the tree shown),
the removed edges that make up $F_O$ \llabel{ref_conceptual2}
are $4 \to 3$, $3 \to 1$, and $3 \to 2$,
and the added edges that make up $F_I$ are $4 \to 1$ and the new edge leading to 2.
The key step in the algorithm is to determine whether the pair of paths
that terminate in a given node in two adjacent trees are mergeable.
The algorithm we use to do this is given as \algorithmref{alg:mergeable},
and works as follows.
If a pair of paths is mergeable,
then the edges of the two paths must lie in $O$ and $I$, respectively.
Suppose an edge in $O$ has child $c$.
To see if $c$ is the base of a pair of mergeable paths,
the algorithm traverses up from $c$ in both $F_O$ and $F_I$;
terminating if a node in the other tree is found
(i.e., if the node traversed in $F_O$ is in $T_{k+1}$ or if the node traversed in $F_I$ is in $T_k$)
or if a pair of traversed nodes have the same time.
If these two traversals end in the same node $p$, the paths are mergeable.
Iterating over all edges in $O$ will thus find all mergeable pairs of paths.
There is often more than one pair of mergeable paths in a tree transition;
so, the algorithm merges pairs of mergeable paths,
starting with pairs that add the smallest number of new edges,
until no more are found.

@@

The main step that is omitted is a description of the \texttt{Merge} operation,
which performs the actual extending of haplotypes.
This algorithm is essentially the same as \texttt{Mergeable} in \algorithmref{alg:mergeable},
except with additional bookkeeping.
Roughly speaking, the algorithm
traverses up from the shared base node $c$,
doing the appropriate operations to insert the nodes along the path found in $T_k$
into the path in $T_{k+1}$.
To do this, some edges that end at $a_k$ will be extended to end at $a_{k+1}$;
some edges that begin at $a_k$ will be postponed to begin at $a_{k+1}$,
and some entirely new edges may be added, as in \figref{fig:extending_diagram}.
Furthermore, the trees $T_k$ and $T_{k+1}$ (and corresponding forests $F_O$ and $F_I$)
need to be updated.

\begin{algorithm2e}[!ht]
	\SetStartEndCondition{ }{}{}%
	\SetKwProg{Fn}{def}{\string:}{}
	\SetKwFunction{Range}{range}%%
	\SetKw{KwTo}{in}
	\SetKwFor{For}{for}{\string:}{}%
	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
	\SetKwFor{While}{while}{:}{fintq}%
	\SetKw{Break}{break}
	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
    \caption{
        Extend haplotypes. Given an ARG $\T$ with $N$ trees $T_1, \ldots, T_N$
        for which edges $O_k$ are removed to transition from $T_k$ to $T_{k+1}$,
        identify and merge all mergeable paths (see text).
        Each child node $c_e$ of each removed edge $e$ is checked to see if it
        is at the base of two mergeable paths;
        paths that add fewer new edges are merged first
        because \texttt{Mergeable} returns the number of new edges required
        to merge the two paths, and $M$ is always less than or equal to
        the minimum number of new edges across all nodes in the current tree transition.
    }
	\label{alg:extend}
	\SetKwFunction{mergeable}{Mergeable}
	\SetKwFunction{merge}{Merge}
	\SetKwFunction{extend}{ExtendHaplotypes}
	\DontPrintSemicolon
    \Fn{\extend($\T$)}{
    	\tcc{For pairs of trees in the ARG, find mergeable paths.}
        \For{$k$ in $1 \ldots N$}{
            \text{Set } $M = 0$ \text{ and }
            $M' = \infty$.\;
            \While{$M < \infty$}{
                \For{$e \in O_k$}{
                    \text{Set } $m_e$ = \mergeable($c_e$, $T_k$, $T_{k+1}$).\tcc*[r]{Count the number of new edges to be merged.}
                    \eIf{$m_e \le M$}{
                        \merge($c_e$, $T_k$, $T_{k+1}$)\tcc*[r]{Merge extended haplotypes from $T_k$ into $T_{k+1}$.}
                    }{
                        \text{Set } $M' = \min(m_e, M)$.\tcc*[r]{Bookkeep $M$ to determine an ordering for path merging.}
                    }
                }
                \text{Set } $M = M'$ \text{ and }
                $M' = \infty$ .\;
            }
        }
    }
\end{algorithm2e}
\revpoint{3}{5}
\revpoint{3}{4}
\revpoint{3}{6}
\begin{algorithm2e}[!ht]
	\SetStartEndCondition{ }{}{}%
	\SetKwProg{Fn}{def}{\string:}{}
	\SetKwFunction{Range}{range}%%
	\SetKw{KwTo}{in}
	\SetKwFor{For}{for}{\string:}{}%
	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
	\SetKwFor{While}{while}{:}{fintq}%
	\SetKw{Break}{break}
	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
    \caption{

>>>>

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\T}{\mathbb{T}}
\newcommand{\ind}{\mathbf{1}}
\newcommand{\tn}{\textnormal}
\newcommand{\ov}{\overline}
\newcommand{\tskit}{\texttt{tskit}}
\newcommand{\tscompare}{\texttt{tscompare}}
\newcommand{\tsinfer}{\texttt{tsinfer}}
\newcommand{\tsdate}{\texttt{tsdate}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\argmax}{\operatorname{argmax}}
\newcommand{\dis}{\operatorname{dis}}
\newcommand{\match}{\underset{\longrightarrow}{\operatorname{match}}}
\newcommand{\invmatch}{\underset{\longleftarrow}{\operatorname{match}}}
\newcommand{\nmatch}{\underset{\centernot\longrightarrow}{\operatorname{match}}}
\newcommand{\comment}[1]{{\color{violet} \it #1}}
\newcommand{\pluseq}{\mathrel{{+}{=}}}
\newcommand{\arf}{\operatorname{ARF}}
\newcommand{\tpr}{\operatorname{TPR}}

@@

So, given the ARG shown in \figref[A\&B]{fig:conceptual},
it should be possible to extend the ancestral haplotype represented by node 3
to obtain the ARG shown in \figref[C\&D]{fig:conceptual},
thus adding additional information to the ARG.
This might be surprising,
as intuition from phylogenetics suggests we can only infer
information about the branching points in the tree, not intermediate (unary) nodes.
The goal of this paper is to answer:
How can we do this, and how accurate is the resulting inference?
See the discussion for more on what this question assumes,
and the connection to parsimony. \revpoint{1}{7}

@@

We work with the \emph{succinct tree sequence} representation of ARGs (henceforth, ``tree sequence''),
to take advantage of the % extensive, stable, and well-documented set of open-source 
tools available in \tskit{} \citep{tskit}, 
and our terminology and notation follows \citet{ralph2020efficiently}.
For our purposes here,
a tree sequence $\T = (N, E)$ contains a set of \emph{nodes} $N$ 
which represent ancestral segments of genome,
and \emph{edges} $E$ which represent relationships between nodes over different regions of the genome.
Each node $n \in N$ has a \emph{time} $t_n$,
which is the amount of time in the past that the individual who carried that segment of genome lived. \revpoint{3}{13}
Some nodes are \emph{samples}, meaning that they represent genome sequences available as data.
% Typically, samples are exclusively leaves of $\T$, however it is not the case that all samples are leaves.
Each edge $e \in E$ describes inheritance between an ancestor and a descendant.
The ancestor and descendant are represented by the parent node $p_e$ and child node $c_e$ respectively \revpoint{2}{1}
over a segment of genome $[\ell_e, r_e)$.
% (The tree sequence data structure typically includes substantially more information, including genotypes, that we will not use in this paper.)
Suppose that the unique elements of the set of left and right edge endpoints
are $0 = a_0 < a_1 < \cdots < a_{n} = L$, where $L$ is the length of the genome.
Using this information, one can construct the sequence of 
trees $\left(T_1,...,T_{n}\right)$ that describe how the nodes are related to each other along the genome:
each $T_k$ is a tree whose nodes are in $N$
and that describes relationships on the half-open interval $[a_{k-1}, a_k)$.
Nodes represent (portions of) ancestral haplotypes, and so
we will use the terms interchangeably.
Not all nodes appear in each tree,
and we say $n \in T_k$ for a node $n$ if the tree $T_k$ describes at least one parent-child relationship
for node $n$.

@@

% JK note: earlier version said "relatively" - relative to what?
An efficient algorithm to do this is described in \algorithmref{alg:extend}.
% and implemented in \tskit{} as \texttt{extend\_haplotypes}.
The algorithm considers each tree transition from $T_k$ to $T_{k+1}$ in turn, updating its internal state
(which includes possibly modifying $T_k$) as it goes.
Suppose we are at the transition from tree $T_k$ to tree $T_{k+1}$,
which is done by first removing a set of edges $O$
and then adding another set of edges $I$.
$O$ defines a sub-forest $F_O$ of $T_k$,
and $I$ defines a sub-forest $F_I$ of $T_{k+1}$.
In \figref{fig:conceptual} (in the portion of the tree shown),
the removed edges that make up $F_O$ \llabel{ref_conceptual2}
are $4 \to 3$, $3 \to 1$, and $3 \to 2$,
and the added edges that make up $F_I$ are $4 \to 1$ and the new edge leading to 2.
The key step in the algorithm is to determine whether the pair of paths
that terminate in a given node in two adjacent trees are mergeable.
The algorithm we use to do this is given as \algorithmref{alg:mergeable},
and works as follows.
If a pair of paths is mergeable,
then the edges of the two paths must lie in $O$ and $I$, respectively.
Suppose an edge in $O$ has child $c$.
To see if $c$ is the base of a pair of mergeable paths,
the algorithm traverses up from $c$ in both $F_O$ and $F_I$;
the variables $y_i$ and $y_o$ indicate whether the next node upwards in the traversals
(i.e., the next older node, as recall $t_n$ is amount of time in the past that $n$ lived) \llabel{time_reminder}
is in $F_I$ or $F_O$.
The traversals terminate if a node in the other tree is found
(i.e., if the node traversed in $F_O$ is in $T_{k+1}$ or if the node traversed in $F_I$ is in $T_k$)
or if a pair of traversed nodes have the same time.
If these two traversals end in the same node $p$, the paths are mergeable.
Iterating over all edges in $O$ will thus find all mergeable pairs of paths.
There is often more than one pair of mergeable paths in a tree transition;
so, the algorithm merges pairs of mergeable paths,
starting with pairs that add the smallest number of new edges,
until no more are found.

Note that the algorithm could be applied to an undated ARG, \llabel{undated}
with some adjustments.
The algorithm only uses node time for two reasons:
convenience, when iterating jointly up the two paths;
and to avoid illegal conditions: if we tried to merge two paths on which
some $u_i$ had the same time as some $v_j$,
then we would have a parent and child with the same time, which is not allowed.

@@

The main step that is omitted is a description of the \texttt{Merge} operation,
which performs the actual extending of haplotypes.
This algorithm is essentially the same as \texttt{MergeNum} in \algorithmref{alg:mergeable},
except with additional bookkeeping.
Roughly speaking, the algorithm
traverses up from the shared base node $c$,
doing the appropriate operations to insert the nodes along the path found in $T_k$
into the path in $T_{k+1}$.
To do this, some edges that end at $a_k$ will be extended to end at $a_{k+1}$;
some edges that begin at $a_k$ will be postponed to begin at $a_{k+1}$,
and some entirely new edges may be added, as in \figref{fig:extending_diagram}.
Furthermore, the trees $T_k$ and $T_{k+1}$ (and corresponding forests $F_O$ and $F_I$)
need to be updated.

\begin{algorithm2e}[!ht]
	\SetStartEndCondition{ }{}{}%
	\SetKwProg{Fn}{def}{\string:}{}
	\SetKwFunction{Range}{range}%%
	\SetKw{KwTo}{in}
	\SetKwFor{For}{for}{\string:}{}%
	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
	\SetKwFor{While}{while}{:}{fintq}%
	\SetKw{Break}{break}
	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
    \caption{
        Extend haplotypes. Given an ARG $\T$ with $N$ trees $T_1, \ldots, T_N$
        for which edges $O_k$ are removed to transition from $T_k$ to $T_{k+1}$,
        identify and merge all mergeable paths (see text).
        Each child node $c_e$ of each removed edge $e$ is checked to see if it
        is at the base of two mergeable paths;
        paths that add fewer new edges are merged first
        because \texttt{MergeNum} returns the number of new edges required
        to merge the two paths, and $M$ is always less than or equal to
        the minimum number of new edges across all nodes in the current tree transition.
    }
	\label{alg:extend}
	\SetKwFunction{mergeable}{MergeNum}
	\SetKwFunction{merge}{Merge}
	\SetKwFunction{extend}{ExtendHaplotypes}
	\DontPrintSemicolon
    \Fn{\extend($\T$)}{
    	\tcc{For pairs of trees in the ARG, find mergeable paths.}
        \For{$k$ in $1 \ldots N$}{
            \text{Set } $M = 0$ \text{ and }
            $M' = \infty$.\;
            \While{$M < \infty$}{
                \For{$e \in O_k$}{
                    \text{Set } $m_e$ = \mergeable($c_e$, $T_k$, $T_{k+1}$).\tcc*[r]{Find the number of new edges required.}
                    \eIf{$m_e \le M$}{
                        \merge($c_e$, $T_k$, $T_{k+1}$)\tcc*[r]{Extended haplotypes from $T_k$ into $T_{k+1}$.}
                    }{
                        \text{Set } $M' = \min(m_e, M')$.
                    }
                }
                \text{Set } $M = M'$ \text{ and }
                $M' = \infty$ .\;
            }
        }
    }
\end{algorithm2e}
